---
title: "Bios 6301: Assignment 4"
author: "Sarah Lotspeich"
date: "10/17/2016"
output: pdf_document
---

**Grade 49/50** Nice job

Due Tuesday, 27 October, 1:00 PM

$5^{n=day}$ points taken off for each day late.

50 points total.

Submit a single knitr file (named homework4.rmd), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as author to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file homework4.rmd or include author name may result in 5 points taken off.

##Question 1##

15 points

A problem with the Newton-Raphson algorithm is that it needs the derivative $f'$. If the derivative is hard to compute or does not exist, then we can use the secant method, which only requires that the function $f$ is continuous.

Like the Newton-Raphson method, the secant method is based on a linear approximation to the function $f$. Suppose that $f$ has a root at $a$. For this method we assume that we have two current guesses, $x_0$ and $x_1$, for the value of $a$. We will think of $x_0$ as an older guess and we want to replace the pair $x_0$, $x_1$ by the pair $x_1$, $x_2$, where $x_2$ is a new guess.

To find a good new guess $x_2$ we first draw the straight line from $(x_0,f(x_0))$ to $(x_1,f(x_1))$, which is called a secant of the curve $y = f(x)$. Like the tangent, the secant is a linear approximation of the behavior of $y = f(x)$, in the region of the points $x_0$ and $x_1$. As the new guess we will use the x-coordinate $x_2$ of the point at which the secant crosses the x-axis.

The general form of the recurrence equation for the secant method is:

$$x_{i+1} = x_i - f(x_i)\frac{x_i - x_{i-1}}{f(x_i) - f(x_{i-1})}$$

Notice that we no longer need to know $f'$ but in return we have to provide two initial points, $x_0$ and $x_1$.

Write a function that implements the secant algorithm. Validate your program by finding the root of the function $f(x) = \cos(x) - x$. 

```{r, tidy=TRUE}
fx <- function(x)
{
  return(cos(x) - x)
}

secantAlg <- function(x0, tol=1e-8, max_iter=100) {
  stopifnot(tol > 0, max_iter > 0)
  x1 <- x0 + 1e6
  iter <- 0
  while(abs(x1-x0) > tol && iter < max_iter) {
    fx0 <- fx(x0)
    fx1 <- fx(x1)
    x2 <- x1 - (fx1*((x1-x0)/(fx1-fx0)))
    x0 <- x1
    x1 <- x2
    iter <- iter + 1
  }
  if(abs(x0-x1) > tol) 
  {
    x1 <- NULL
    print("Algorithm failed to converge!")
  }
  return(x1)
}

x2 <- secantAlg(10) #find the root of f(x) = cos(x) + x
curve(fx)
abline(h=0) #plot x axis
abline(v=x2, col="red", lty=2) #plot root
```

Compare its performance with the Newton-Raphson method -- which is faster, and by how much? For this example $f'(x) = -\sin(x) - 1$.

```{r, tidy=TRUE}
#Newton-Raphson Method
fxdx <- function(x)
{
  return(-sin(x) - 1)
}

newtonRaphsonAlg <- function(x0, max_iter=100) {
  stopifnot(max_iter > 0)
  iter <- 0
  while(iter < max_iter) {
    fx0 <- fx(x0)
    fxdx0 <- fxdx(x0)
    x1 <- x0 - (fx0/fxdx0)
    x0 <- x1
    iter <- iter + 1
  }
  #if (iter == max_iter)
  #{
    #print("Algorithm failed to converge!")
    #return(NULL)
  #}
  #else 
  return(x0)
}

x1 <- newtonRaphsonAlg(x0=10)
curve(fx)
abline(h=0) #plot x axis
abline(v=x1, col="red", lty=2) #plot root

system.time(replicate(10000,secantAlg(10)))
system.time(replicate(10000,newtonRaphsonAlg(10)))
```

With fewer inputs, the Newton-Raphson Method returned the exact same root for the equation $f(x) = cos(x) + x$. However, this method is limited by our ability to differentiate the function of interest $f(x)$. Overall, the function to carry out the Newton-Raphson algorithm is much simpler than that of the Secant algorithm, because the latter requires two points and a given amount of tolerance that will continue to move toward each other until capturing the root. When repeating each algorithm 10,000 times and clocking them with system.time(), the Secant Algorithm seemed to blow the Newton-Raphson Algorithm out of the water! The Newton-Raphson Algorithm took more than 3 seconds to execute 10,000 times, whereas the Secant Algorithm accomplished the same task in nearly half a second.  

##Question 2##

18 points

The game of craps is played as follows. First, you roll two six-sided dice; let x be the sum of the dice on the first roll. If x = 7 or 11 you win, otherwise you keep rolling until either you get x again, in which case you also win, or until you get a 7 or 11, in which case you lose.

Write a program to simulate a game of craps. You can use the following snippet of code to simulate the roll of two (fair) dice:

    x <- sum(ceiling(6*runif(2)))
    
The instructor should be able to easily import and run your program (function), and obtain output that clearly shows how the game progressed. Set the RNG seed with set.seed(100) and show the output of three games. (lucky 13 points)

```{r, tidy=TRUE}
set.seed(100) #set seed as specified

crapsGame <- function(output)
{
  keepRolling <- TRUE
  firstRoll <- TRUE
  point <- 0
  while(keepRolling)
  {
  x <- sum(ceiling(6*runif(2))) #simulate dice roll
    if (firstRoll == TRUE)
    {
      if (output == TRUE)
      {
        print("New game.")  
        print(paste("Sum of the first roll was",x)) #print first x value
        if (x == 7 || x == 11)
        {
            print("Win!")
            return(print("End of game."))
        }
        else
        {
            print(paste("Roll again with ", point <- x, " as point."))
            firstRoll <- FALSE 
        }
      }
      else
      {
        if (x == 7 || x == 11)
        {
          return(1)
        }
        else
        {
          point <- x
          firstRoll <- FALSE 
        } 
      }
    }
    else
    {
      if (output == TRUE)
      {
        print(paste("Next sum of roll was ",x))
        if (x == point)
        {
          print("Win!")
          return(print("End of game."))
        }
        if (x == 7 || x == 11)
        {
          print("Lose!")
          return(print("End of game."))
        }
      }
      else
      {
        if (x == point)
        {
          return(1)
        }
        if (x == 7 || x == 11)
        {
          return(0)
        }
      }
    }
  }
}

for (i in 1:3)
{
  crapsGame(TRUE)
}
```

Find a seed that will win ten straight games. Consider adding an argument to your function that disables output. Show the output of the ten games. (5 points)

```{r, tidy=TRUE}
seeds <- seq(1,10000)

trySeed <- function(seed)
{
  set.seed(seed)
  wins <- NULL
  for (i in 1:10)
  {
    wins[i] <- crapsGame(FALSE)
  }
  return(sum(wins)) #return wins for that seed
}

winsVector <- sapply(seeds,trySeed)
compSeeds <- as.data.frame(cbind(seeds,winsVector))
which(compSeeds$winsVector == 10) #find seeds between 1 and 10,000 that will yield 10 wins in a row

set.seed(880) #display output for first seed with 10 straight wins
for (i in 1:10)
{
  crapsGame(TRUE)
}
```

##Question 3##

12 points

Obtain a copy of the football-values lecture. Save the five 2015 CSV files in your working directory.

I 

```{r, tidy=TRUE}
#knitr::opts_knit$set(root.dir = normalizePath("~/Dropbox/Vanderbilt/Intro to Statistical Computing/football-values-master"))
knitr::opts_knit$set(root.dir = normalizePath("."))
year <- 2016
positions <- c('k','qb','rb','te','wr')
csvfile <- paste('proj_', positions, substr(year, 3, 4), '.csv', sep='')
files <- file.path(year, csvfile)
names(files) <- positions
```

Modify the code to create a function. This function will create dollar values given information (as arguments) about a league setup. It will return a data.frame and write this data.frame to a CSV file. The final data.frame should contain the columns 'PlayerName', 'pos', 'points', 'value' and be ordered by value descendingly. Do not round dollar values.

Note that the returned data.frame should have sum(posReq)*nTeams rows.

Define the function as such (6 points):

```{r, tidy=TRUE}
# path: directory path to input files
# file: name of the output file; it should be written to path
# nTeams: number of teams in league
# cap: money available to each team
# posReq: number of starters for each position
# points: point allocation for each category
#knitr::opts_knit$set(root.dir = normalizePath("~/Dropbox/Vanderbilt/Intro to Statistical Computing/football-values-master"))
knitr::opts_knit$set(root.dir = normalizePath("."))

ffvalues <- function(path, file='outfile.csv', nTeams=12, cap=200, posReq=c(qb=1, rb=2, wr=3, te=1, k=1),points=c(fg=4, xpt=1, pass_yds=1/25, pass_tds=4, pass_ints=-2, rush_yds=1/10, rush_tds=6, fumbles=-2, rec_yds=1/20, rec_tds=6)) 
{
  ## read in CSV files
  knitr::opts_chunk$set(root.dir = normalizePath(path))
  year <- 2016
  positions <- c('k','qb','rb','te','wr')
  csvfile <- paste('proj_', positions, substr(year, 3, 4), '.csv', sep='')
  files <- file.path(year, csvfile)
  names(files) <- positions
  k <- read.csv(files['k'], header=TRUE, stringsAsFactors=FALSE)
  qb <- read.csv(files['qb'], header=TRUE, stringsAsFactors=FALSE)
  rb <- read.csv(files['rb'], header=TRUE, stringsAsFactors=FALSE)
  te <- read.csv(files['te'], header=TRUE, stringsAsFactors=FALSE)
  wr <- read.csv(files['wr'], header=TRUE, stringsAsFactors=FALSE)
  
  # generate unique list of column names
  cols <- unique(c(names(k), names(qb), names(rb), names(te), names(wr)))

  # create a new column in each data.frame
  k[,'pos'] <- 'k'
  qb[,'pos'] <- 'qb'
  rb[,'pos'] <- 'rb'
  te[,'pos'] <- 'te'
  wr[,'pos'] <- 'wr'

  # append 'pos' to unique column list
  cols <- c(cols, 'pos')
  
  # create common columns in each data.frame
  # initialize values to zero
  k[,setdiff(cols, names(k))] <- 0
  qb[,setdiff(cols, names(qb))] <- 0
  rb[,setdiff(cols, names(rb))] <- 0
  te[,setdiff(cols, names(te))] <- 0
  wr[,setdiff(cols, names(wr))] <- 0

  # combine data.frames by row, using consistent column order
  x <- rbind(k[,cols], qb[,cols], rb[,cols], te[,cols], wr[,cols])
  
  ## calculate dollar values
  # convert NFL stat to fantasy points
  x[,'p_fg'] <- x[,'fg']*points[1] 
  x[,'p_xpt'] <- x[,'xpt']*points[2]
  x[,'p_pass_yds'] <- x[,'pass_yds']*points[3]
  x[,'p_pass_tds'] <- x[,'pass_tds']*points[4]
  x[,'p_pass_ints'] <- x[,'pass_ints']*points[5]
  x[,'p_rush_yds'] <- x[,'rush_yds']*points[6]
  x[,'p_rush_tds'] <- x[,'rush_tds']*points[7]
  x[,'p_fumbles'] <- x[,'fumbles']*points[8]
  x[,'p_rec_yds'] <- x[,'rec_yds']*points[9]
  x[,'p_rec_tds'] <- x[,'rec_tds']*points[10]

  x[,'points'] <- rowSums(x[,grep("^p_", names(x))]) 
  
  # create new data.frame ordered by points descendingly
  x2 <- x[order(x[,'points'], decreasing=TRUE),]

  # determine the row indeces for each position
  k.ix <- which(x2[,'pos']=='k')
  qb.ix <- which(x2[,'pos']=='qb')
  rb.ix <- which(x2[,'pos']=='rb')
  te.ix <- which(x2[,'pos']=='te')
  wr.ix <- which(x2[,'pos']=='wr')

  # calculate marginal points by subtracting "baseline" player's points
  x2[k.ix, 'marg'] <- x2[k.ix,'points'] - ifelse(length(x2[k.ix[(posReq["k"]*nTeams)],'points'])==0,0,x2[k.ix[(posReq["k"]*nTeams)],'points'])
  x2[qb.ix, 'marg'] <- x2[qb.ix,'points'] - x2[qb.ix[(posReq["qb"]*nTeams)],'points']
  x2[rb.ix, 'marg'] <- x2[rb.ix,'points'] - x2[rb.ix[(posReq["rb"]*nTeams)],'points']
  x2[te.ix, 'marg'] <- x2[te.ix,'points'] - x2[te.ix[(posReq["te"]*nTeams)],'points']
  x2[wr.ix, 'marg'] <- x2[wr.ix,'points'] - x2[wr.ix[(posReq["wr"]*nTeams)],'points']

  # create a new data.frame subset by non-negative marginal points
  x3 <- x2[x2[,'marg'] >= 0,]

  # re-order by marginal points
  x3 <- x3[order(x3[,'marg'], decreasing=TRUE),]

  # reset the row names
  rownames(x3) <- NULL

  # calculation for player value
  x3[,'value'] <- x3[,'marg']*(12*200-nrow(x3))/sum(x3[,'marg']) + 1
  x3[,'value'] <- x3[,'marg']*(nTeams*cap-nrow(x3))/sum(x3[,'marg']) + 1

  # create a data.frame with more interesting columns
  x4 <- x3[,c('PlayerName','pos','points','value')]
  ## save dollar values as CSV file
  write.csv(x4, file=file)
  ## return data.frame with dollar values
  return(x4)
}

x1 <- ffvalues(".")
```

How many players are worth more than $20? (1 point)

```{r}
nrow(x1[x1[,'value'] > 20,])
```

Who is 15th most valuable running back (rb)? (1 point)

```{r}
x1[which(x1[,'pos']=='rb')[15],"PlayerName"]
```

Call: 

```{r}
x2 <- ffvalues(getwd(), '16team.csv', nTeams=16, cap=150)
```

How many players are worth more than $20? (1 point)

```{r}
nrow(x2[x2[,'value'] > 20,])
```

How many wide receivers (wr) are in the top 40? (1 point)

```{r}
top40 <- x2[1:40,]
length(which(top40[,'pos']=='wr'))
```

Call:

```{r, tidy=TRUE}
x3 <- ffvalues('.', 'qbheavy.csv', posReq=c(qb=2, rb=2, wr=3, te=1, k=0),
        points=c(fg=0, xpt=0, pass_yds=1/25, pass_tds=6, pass_ints=-2,
                rush_yds=1/10, rush_tds=6, fumbles=-2, rec_yds=1/20, rec_tds=6))
```

How many players are worth more than $20? (1 point)

```{r}
nrow(x3[x3[,'value'] > 20,])
```

**JC Grading -1**

**Comment:** There are actually 51 players.  Look to your ifelse statement on assigning marginal worth: ifelse(length(x2[k.ix[(posReq["k"]*nTeams)],'points'])==0,0,x2[k.ix[(posReq["k"]*nTeams)],'points']).  This assigns a value of 0.  Assign a negative value, then it'll be appropriately removed when calling the condition: x3 <- x2[x2[,'marg'] >= 0,]


How many quarterbacks (qb) are in the top 30? (1 point)

```{r}
top30 <- x3[1:30,]
length(which(top30[,'pos']=='qb'))
```

##Question 4##

5 points

This code makes a list of all functions in the base package:

```{r}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)
```

Using this list, write code to answer these questions.

Which function has the most arguments? (3 points)

```{r, tidy=TRUE}
getArgs <- function(x)
{
  parameters <- formals(x) #get the parameters for the function x
  return(length(parameters)) #find length of the list of the parameters
}

numParameters <- sapply(funs, getArgs) #find the number of parameters for all of the functions in funs
numParameters <- sort(numParameters)
mostArgs <- which.max(numParameters)
numParameters[mostArgs]
```

How many functions have no arguments? (2 points)

```{r}
length(numParameters[numParameters == 0])
```

Hint: find a function that returns the arguments for a given function.